---
title: "Is the gender gap in AI closing? Evidence from Wikipedia"
author: "Naomi, Amber, Tristan (Group 73)"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = getwd())   # lets the section files find data/ and figures/
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(3020)
dir.create("figures", showWarnings = FALSE)
```

```{r load-data, include=FALSE}
people = read.csv("data/people.csv", as.is = TRUE)
edges  = read.csv("data/edges.csv", as.is = TRUE)

# birth year 2000 = Wikidata's placeholder for "20th century", so treat it as unknown
people$birth_year[people$birth_year == 2000] = NA

people$era = ifelse(is.na(people$birth_year), "unknown",
             ifelse(people$birth_year < 1950, "pre-1950",
             ifelse(people$birth_year < 1970, "1950-69",
             ifelse(people$birth_year < 1990, "1970-89", "1990+"))))
people$recognition_lag = people$created_year - people$birth_year

# women and men with a known birth year
known = people[people$era != "unknown" & people$gender_group %in% c("female", "male"), ]
known$era = factor(known$era, levels = c("pre-1950", "1950-69", "1970-89", "1990+"))
```

# Introduction

Women remain under-represented in artificial intelligence (AI). This report uses
Wikipedia to ask whether that gap is closing. We study `r nrow(people)` AI
researchers with an English Wikipedia article and answer five research questions:

1. Has the share of women in AI increased across birth eras?
2. Are women recognised on Wikipedia later than men?
3. Has the language used to describe women and men changed over time?
4. Which AI subfields emerge from the biography text, and what share of each is women?
5. Are women central in the network of biographies that link to each other?

Code and data: https://github.com/minhhien365/COMP-3020---Group-Assignment

# Data

```{r child = "sections/01_data_collection.Rmd"}
```

# Analysis

```{r child = "sections/02_rq1_rq2.Rmd"}
```

```{r child = "sections/03_rq3_rq4.Rmd"}
```

```{r child = "sections/04_rq5.Rmd"}
```

# Overall findings and limitations

```{r child = "sections/05_overall_findings.Rmd"}
```
