## Data collection

### How the data was accessed

We collected all the data from three free Wikimedia APIs. None of them need an API key.

1. **Wikidata (SPARQL):** we used this to find the AI researchers, along with their
   gender, birth year and a short description.
2. **Wikipedia API:** we used this to get each person's article intro, article length,
   the date the article was created, and the links to other articles.
3. **Wikimedia page views API:** we used this to get how many times each article was
   viewed.

All three APIs return JSON, so we read them with `fromJSON()` from `jsonlite`, like
in tut 2. Wikimedia blocks programs that don't identify themselves or that send
requests too fast, so we wrote a helper function that sends a User-Agent and retries
a request up to three times. We also paused between requests.

The code below is not run when the report is knitted, because it takes about 25
minutes and needs the internet. We saved the results to `data/people.csv` and
`data/edges.csv` and the report loads those instead.

```{r dc-setup, eval=FALSE}
library("jsonlite")

UA = "COMP3020-group-project (student project)"   # identifies our program to Wikimedia

# download a URL and turn the JSON into R objects; retry up to 3 times
get_json = function(u) {
  for (attempt in 1:3) {
    result = tryCatch({
      con = url(u, headers = c("User-Agent" = UA))
      txt = paste(readLines(con, warn = FALSE, encoding = "UTF-8"), collapse = "")
      close(con)
      fromJSON(txt)
    }, error = function(e) NULL)
    if (!is.null(result)) return(result)
    Sys.sleep(2 * attempt)   # wait longer before each retry
  }
  return(NULL)
}

WIKI = "https://en.wikipedia.org/w/api.php?format=json&formatversion=2"
```

### Step 1: finding the AI researchers (Wikidata)

We kept every human with an English Wikipedia article whose field of work is
artificial intelligence (or a sub-field of it, like machine learning), or whose
occupation is AI researcher. The `/wdt:P279*` part of the query is what pulls in
the sub-fields.

```{r dc-wikidata, eval=FALSE}
query = '
SELECT ?person ?article ?genderLabel ?description
       (MIN(YEAR(?dob)) AS ?birth_year) (SAMPLE(?sl) AS ?sitelinks)
WHERE {
  { ?person wdt:P101/wdt:P279* wd:Q11660 . }
  UNION
  { ?person wdt:P106/wdt:P279* wd:Q15976092 . }
  ?person wdt:P31 wd:Q5 ;
          wdt:P21 ?gender ;
          wikibase:sitelinks ?sl .
  OPTIONAL { ?person wdt:P569 ?dob . }
  OPTIONAL { ?person schema:description ?description .
             FILTER(LANG(?description) = "en") }
  ?article schema:about ?person ;
           schema:isPartOf <https://en.wikipedia.org/> .
  ?gender rdfs:label ?genderLabel . FILTER(LANG(?genderLabel) = "en")
}
GROUP BY ?person ?article ?genderLabel ?description
'
wd = get_json(paste0("https://query.wikidata.org/sparql?format=json&query=",
                     URLencode(query, reserved = TRUE)))
b = wd$results$bindings

people = data.frame(
  qid         = sub(".*/", "", b$person$value),
  title       = gsub("_", " ", URLdecode(sub(".*/wiki/", "", b$article$value))),
  gender      = b$genderLabel$value,
  description = b$description$value,
  birth_year  = as.numeric(b$birth_year$value),
  sitelinks   = as.numeric(b$sitelinks$value),
  stringsAsFactors = FALSE)
people = people[!duplicated(people$qid), ]

people$gender_group = ifelse(people$gender %in% c("female", "male"),
                             people$gender, "other/non-binary")
```

### Step 2: article text, length and creation date (Wikipedia API)

We got the intro text and article length 20 articles at a time. For the creation
date we asked for each article's first edit, which has to be done one article at a
time. Recognition lag (for RQ2) is then the creation year minus the birth year.

```{r dc-wikipedia, eval=FALSE}
# intro text and length, 20 articles per request
groups = split(1:nrow(people), ceiling(seq_len(nrow(people)) / 20))
for (g in groups) {
  u = paste0(WIKI, "&action=query&prop=extracts|info&exintro=1&explaintext=1&exlimit=20",
             "&titles=", URLencode(paste(people$title[g], collapse = "|"), reserved = TRUE))
  d = get_json(u)
  if (is.null(d)) next
  m = match(d$query$pages$title, people$title)
  people$intro_text[m[!is.na(m)]]    = d$query$pages$extract[!is.na(m)]
  people$article_bytes[m[!is.na(m)]] = d$query$pages$length[!is.na(m)]
  Sys.sleep(0.5)
}

# creation date = timestamp of the first edit
for (i in 1:nrow(people)) {
  u = paste0(WIKI, "&action=query&prop=revisions&rvprop=timestamp&rvlimit=1&rvdir=newer",
             "&titles=", URLencode(people$title[i], reserved = TRUE))
  d = get_json(u)
  if (!is.null(d$query$pages$revisions[[1]]))
    people$created[i] = d$query$pages$revisions[[1]]$timestamp
  Sys.sleep(0.2)
}
people$created_year    = as.numeric(substr(people$created, 1, 4))
people$recognition_lag = people$created_year - people$birth_year
```

### Step 3: page views (Wikimedia page views API)

We added up the monthly views from October 2025 to September 2026. We only counted
human readers, so bot traffic is left out.

```{r dc-pageviews, eval=FALSE}
for (i in 1:nrow(people)) {
  t = URLencode(gsub(" ", "_", people$title[i]), reserved = TRUE)
  u = paste0("https://wikimedia.org/api/rest_v1/metrics/pageviews/per-article/",
             "en.wikipedia/all-access/user/", t, "/monthly/20251001/20260930")
  d = get_json(u)
  people$pageviews_12m[i] = if (is.null(d)) 0 else sum(d$items$views)
  Sys.sleep(0.2)
}
```

### Step 4: links between researchers (Wikipedia API)

For each person we got every article their page links to and only kept the links
that go to someone else in our dataset. The API only gives 500 links per request,
so for long articles we kept asking for the next batch until there were none left.

```{r dc-links, eval=FALSE}
edges = data.frame(from = character(), to = character(), stringsAsFactors = FALSE)
for (i in 1:nrow(people)) {
  linked = c(); cont = ""
  repeat {
    u = paste0(WIKI, "&action=query&generator=links&gplnamespace=0&gpllimit=max&redirects=1",
               "&titles=", URLencode(people$title[i], reserved = TRUE), cont)
    d = get_json(u)
    if (is.null(d)) break
    linked = c(linked, d$query$pages$title)
    if (is.null(d$continue$gplcontinue)) break
    cont = paste0("&gplcontinue=", URLencode(d$continue$gplcontinue, reserved = TRUE))
  }
  targets = setdiff(people$qid[people$title %in% linked], people$qid[i])  # no self-links
  if (length(targets) > 0) edges = rbind(edges, data.frame(from = people$qid[i], to = targets))
  Sys.sleep(0.2)
}
edges = unique(edges)

write.csv(people, "data/people.csv", row.names = FALSE, fileEncoding = "UTF-8")
write.csv(edges,  "data/edges.csv",  row.names = FALSE, fileEncoding = "UTF-8")
```

### R packages

- `jsonlite` for reading the API data
- base R for the hypothesis tests and plots (RQ1, RQ2)
- `tm`, `SnowballC` and `wordcloud` for the text analysis (RQ3, RQ4)
- `igraph` for the network analysis (RQ5)

### Variables and volume

```{r dc-summary}
table(people$gender_group)
```

| Variable | Description |
|:-------------------------|:-------------------------------------------------------|
| `qid`, `title` | Wikidata ID and Wikipedia article title |
| `gender_group` | female, male or other/non-binary |
| `birth_year`, `era` | birth year and birth era (pre-1950, 1950-69, 1970-89, 1990+) |
| `created_year` | year the Wikipedia article was created |
| `recognition_lag` | `created_year - birth_year` (RQ2) |
| `intro_text` | the article's intro text (RQ3, RQ4) |
| `article_bytes`, `pageviews_12m`, `sitelinks` | article length, page views in the last year, number of language versions |
| `in_degree`, `out_degree` | links to / from other people in the dataset (RQ5) |

### Data cleaning

When Wikidata only knows someone was born in the 20th century, it saves their birth
year as 2000. This happened for 21 people (for example Alex Krizhevsky and Ruslan
Salakhutdinov), so we set their birth year to unknown. That leaves
`r sum(is.na(people$birth_year))` people
(`r round(100 * mean(is.na(people$birth_year)))`%) with no birth year. They are left
out of anything split by era, which leaves `r nrow(known)` women and men.

### Limitations

- Not every AI researcher is tagged with a field of work on Wikidata, so we only
  have the ones that are tagged.
- We only have people who already have an English Wikipedia article. Anyone
  Wikipedia has missed completely can't be in the data, and this could affect women
  more.
- Birth year is missing more often for women
  (`r round(100 * mean(is.na(people$birth_year[people$gender_group == "female"])))`%)
  than men
  (`r round(100 * mean(is.na(people$birth_year[people$gender_group == "male"])))`%),
  so women are under-counted in the era analyses.
- Gender on Wikidata is almost all female or male. Only
  `r sum(!people$gender_group %in% c("female", "male"))` people are recorded as
  something else, which is too few to analyse.
- Wikipedia and Wikidata change every day, so running the code again would give
  slightly different numbers.
- If a request still failed after three tries we skipped it, so a few values (like
  someone's links) might be missing.
