---
editor_options:
  markdown:
    wrap: 72
output: pdf_document
---

## RQ1: Has the share of women in AI increased across birth eras?

**Hypotheses.**

- H0: Gender and birth era are independent.
- HA: Gender and birth era are not independent.

**Test choice.** Both variables are categorical, so we use a chi-squared
test for independence on the era x gender table of counts. The built-in
test needs every expected count to be greater than 5, so we check this
first.

We first count the women and men in each birth era, convert these counts
to percentages, and check the expected counts.

```{r rq1-table}
# writes p-values as "< 0.001" or "= 0.94" in the text
fmt_p = function(p) if (p < 0.001) "< 0.001" else paste("=", signif(p, 2))

tab = table(known$era, known$gender_group)
tab

# share of women and men in each era (%)
round(prop.table(tab, 1) * 100, 1)

# assumption check: expected counts
ct = chisq.test(tab)
round(ct$expected, 1)
```

Women are a minority in every era. All expected counts are above 5
except women born 1990+ (`r round(ct$expected["1990+", "female"], 1)`),
because that era has only `r sum(tab["1990+", ])` people. So we need
simulation.

```{r rq1-test}
ct = chisq.test(tab, simulate.p.value = TRUE)
ct
```

The p-value is `r fmt_p(ct$p.value)`, so we reject H0: the share of
women differs between eras. The test does not show *how* it differs, so
we plot the share of women in each era below. The error bars are
bootstrap 95% confidence intervals; wider bars mean less certainty,
which happens when an era has fewer people.

```{r rq1-plot, fig.cap="Share of women among AI researchers by birth era. Error bars are bootstrap 95% confidence intervals."}
# share of women in each era (%) and number of people in each era
share = prop.table(tab, 1)[, "female"] * 100
n = rowSums(tab)

# bootstrap 95% confidence interval for each era
lower = c()
upper = c()
for (e in levels(known$era)) {
  women = known$gender_group[known$era == e] == "female"       # TRUE/FALSE for each person
  boot = replicate(1000, mean(sample(women, replace = TRUE))) * 100
  lower = c(lower, quantile(boot, 0.025))
  upper = c(upper, quantile(boot, 0.975))
}

# line chart with error bars
plot(1:4, share, type = "b", pch = 19, col = "darkorange", ylim = c(0, 40), xaxt = "n",
     main = "Share of women in AI by birth era", xlab = "Birth era", ylab = "Women (%)")
axis(1, at = 1:4, labels = paste0(names(share), " (n=", n, ")"))
arrows(1:4, lower, 1:4, upper, angle = 90, code = 3, length = 0.05)

# save a copy for the poster
dev.copy(png, file.path(fig_dir, "rq1_share_women_by_era.png"), width = 1400, height = 900, res = 200)
invisible(dev.off())
```

The share of women rises from `r round(share[1])`% (pre-1950) to
`r round(share[2])`% (1950-69) and `r round(share[3])`% (1970-89), and
the error bars for pre-1950 (`r round(lower[1])`-`r round(upper[1])`%)
and 1970-89 (`r round(lower[3])`-`r round(upper[3])`%) do not overlap.
The drop to `r round(share[4])`% for 1990+ rests on only `r n[4]` people
(`r tab["1990+", "female"]` women), so its error bar is very wide
(`r round(lower[4])`-`r round(upper[4])`%) and it should not be read as
a reversal; younger researchers have also had less time to become
notable.

**Answer to RQ1:** yes, the share of women in AI has increased across
birth eras, although women remain a minority in every era.

## RQ2: Are women recognised on Wikipedia later than men?

To measure how quickly a person is recognised on Wikipedia, we use
recognition lag = year the article was created - birth year. For
example, someone born in 1960 whose article was created in 2010 has a
recognition lag of 50 years. A shorter lag means the person was
recognised earlier in their life. If women in AI are overlooked, we
would expect them to receive an article later in life than men, so their
recognition lag would be longer.

Recognition lag has two important limitations. First, Wikipedia was only
launched in 2001, so anyone born long before then automatically has a
big lag, whatever their gender. This means lag depends strongly on birth
era, so we also compare women and men within each era. Second, the
creation date reflects when a Wikipedia editor chose to write the
article, not when the person achieved something notable. This is exactly
why it is useful here: it measures when Wikipedia recognised them, not
their career itself.

**Hypotheses.**

- H0: There is no difference in mean recognition lag between women and
  men.
- HA: There is a difference in mean recognition lag between women and
  men.

**Test choice.** We use a randomisation test for the difference in means
and a Wilcoxon rank-sum test as a check.

We first separate the recognition lags of women and men and compare
their size, mean and median.

```{r rq2-data}
lagF = known$recognition_lag[known$gender_group == "female"]
lagM = known$recognition_lag[known$gender_group == "male"]

data.frame(group  = c("Women", "Men"),
           n      = c(length(lagF), length(lagM)),
           mean   = round(c(mean(lagF), mean(lagM)), 1),
           median = c(median(lagF), median(lagM)))
```

Women and men have almost identical recognition lags overall: a mean of
`r round(mean(lagF), 1)` vs `r round(mean(lagM), 1)` years, and a median
of `r median(lagF)` vs `r median(lagM)` years. There are far more men
(`r length(lagM)`) than women (`r length(lagF)`) with a known birth
year.

To test whether this difference could have occurred by chance, we run
the randomisation test: if H0 is true, the gender labels do not matter,
so we shuffle them 2000 times and recalculate the difference in means
each time.

```{r rq2-rand, fig.height=3.5, fig.cap="Randomisation distribution of the difference in mean recognition lag (women minus men) under H0. The red line is the observed difference."}
obs_diff = mean(lagF) - mean(lagM)   # women minus men
combined = c(lagF, lagM)
nF = length(lagF)

rand_diff = replicate(2000, {
  shuffled = sample(combined)
  mean(shuffled[1:nF]) - mean(shuffled[-(1:nF)])
})

# two-sided p-value: proportion of random differences at least as extreme as observed
p_rand = mean(abs(rand_diff) >= abs(obs_diff))
c(observed_difference = round(obs_diff, 2), p_value = p_rand)

hist(rand_diff, breaks = 40, col = "grey80", main = "",
     xlab = "Difference in sample means (women - men, years)")
abline(v = obs_diff, col = "red", lwd = 2)
```

The observed difference in means is only `r round(obs_diff, 2)` years.
In the histogram, the red line sits in the middle of the randomisation
distribution, so a difference this small easily happens by chance. The
p-value is `r p_rand`, so we do not reject H0: overall there is no
evidence that women's lag differs from men's.

Recognition lag is skewed, with a few people having very long lags that
can pull the mean up or down. As a check, we also run a Wilcoxon
rank-sum test, which compares the ranks of the lags instead of their
actual values.

```{r rq2-wilcox}
wt = wilcox.test(lagF, lagM)
wt
```

The Wilcoxon test agrees: W =
`r format(unname(wt$statistic), scientific = FALSE)`, p
`r fmt_p(wt$p.value)`, so there is no overall difference.

Because lag depends strongly on birth era, we repeat the Wilcoxon test
separately within each era to compare like with like.

```{r rq2-era}
era_tab = data.frame(era = levels(known$era), median_women = NA, median_men = NA, p_value = NA)

for (i in 1:4) {
  f = known$recognition_lag[known$era == era_tab$era[i] & known$gender_group == "female"]
  m = known$recognition_lag[known$era == era_tab$era[i] & known$gender_group == "male"]
  era_tab$median_women[i] = median(f)
  era_tab$median_men[i]   = median(m)
  era_tab$p_value[i]      = wilcox.test(f, m, exact = FALSE)$p.value
}

era_show = era_tab
era_show$p_value = format.pval(era_tab$p_value, digits = 2, eps = 0.001)
knitr::kable(era_show, digits = 1,
             caption = "Median recognition lag (years) by gender within each birth era, with Wilcoxon rank-sum p-values.")
```

Splitting by era changes the picture. Women born before 1950 waited
longer than men (median `r era_tab$median_women[1]` vs
`r era_tab$median_men[1]` years, p `r fmt_p(era_tab$p_value[1])`), and
so did women born 1950-69 (`r era_tab$median_women[2]` vs
`r era_tab$median_men[2]` years, p `r fmt_p(era_tab$p_value[2])`). For
1970-89 (`r era_tab$median_women[3]` vs `r era_tab$median_men[3]`, p
`r fmt_p(era_tab$p_value[3])`) and 1990+ (`r era_tab$median_women[4]` vs
`r era_tab$median_men[4]`, p `r fmt_p(era_tab$p_value[4])`) there is no
significant difference.

Finally, we plot the lags of women and men side by side within each era,
to see both the effect of era and any gender gap within it.

```{r rq2-plot, fig.width=8, fig.cap="Recognition lag (years from birth to Wikipedia article) by gender and birth era."}
boxplot(recognition_lag ~ gender_group + era, data = known,
        col = c("pink", "lightblue"), at = c(1, 2, 4, 5, 7, 8, 10, 11), xaxt = "n",
        main = "Recognition lag by gender and birth era",
        xlab = "Birth era", ylab = "Years from birth to Wikipedia article")
axis(1, at = c(1.5, 4.5, 7.5, 10.5), labels = levels(known$era))
legend("topright", legend = c("Female", "Male"), fill = c("pink", "lightblue"))

# save a copy for the poster
dev.copy(png, file.path(fig_dir, "rq2_lag_by_gender_era.png"), width = 1600, height = 900, res = 200)
invisible(dev.off())
```

The boxplot shows lag falling sharply from older to younger eras for
both genders, mostly because Wikipedia only began in 2001. In the two
oldest eras the women's boxes sit higher than the men's, while from 1970
onwards they overlap almost completely.

**Answer to RQ2:** overall, women are not recognised later than men.
However, women born before 1970 received their article later in life
than comparable men. The overall test hides this because women are
concentrated in the younger eras, which have shorter lags. The gap has
disappeared for recent generations.
