---
title: "RQ5: Are women central in the network of AI biographies?"
output: pdf_document
---

```{r rq5-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
people = read.csv("data/people.csv", as.is = TRUE, encoding = "UTF-8")
edges  = read.csv("data/edges.csv",  as.is = TRUE)

# same birth year fix as report.Rmd
people$birth_year[people$birth_year == 2000] = NA
people$era = ifelse(is.na(people$birth_year), "unknown",
             ifelse(people$birth_year < 1950, "pre-1950",
             ifelse(people$birth_year < 1970, "1950-69",
             ifelse(people$birth_year < 1990, "1970-89", "1990+"))))
```

## The link network

Nodes are the 919 AI researchers. There is an edge from A to B when A's English Wikipedia article links to B's article, keeping only links to other people in the dataset. The graph is directed, because a link on A's page does not mean B's page links back, and PageRank needs that direction.

```{r rq5-graph}
library(igraph)

# every edge points to someone in the dataset, with no self-links or duplicates
c(in_dataset = all(c(edges$from, edges$to) %in% people$qid),
  self_links = sum(edges$from == edges$to),
  duplicates = sum(duplicated(edges)))

g = graph_from_data_frame(edges, directed = TRUE,
      vertices = people[, c("qid", "title", "gender_group", "era")])
summary(g)

# share of links that go both ways (A -> B and B -> A)
mean(paste(edges$to, edges$from) %in% paste(edges$from, edges$to))
```

The network has 919 nodes and 2,204 directed edges. 78% of the links go both ways, mostly because of Wikipedia navigation boxes (navboxes), which link everyone on a list to everyone else.

## Network properties

```{r rq5-properties}
comp_w = components(g, mode = "weak")
lcc = induced_subgraph(g, which(comp_w$membership == which.max(comp_w$csize)))

data.frame(density         = round(edge_density(g), 4),
           isolates        = sum(degree(g) == 0),
           weak_components = comp_w$no,
           largest_comp    = vcount(lcc),
           lcc_edges       = ecount(lcc),
           lcc_diameter    = diameter(lcc, directed = FALSE),
           lcc_mean_dist   = round(mean_distance(lcc, directed = FALSE), 2))
```

The network is very sparse (density 0.0026). 474 researchers (52%) have no links at all. Most of the rest form one large weakly connected component of 365 people that holds 97% of the links, with a diameter of 14 and an average distance of 4.3 links. Centrality is calculated on the full graph, and the largest component is only used for distances and the network plot.

```{r rq5-isolated}
iso_tab = table(V(g)$gender_group, ifelse(degree(g) > 0, "connected", "isolated"))
round(prop.table(iso_tab, 1), 2)

ct_iso = chisq.test(iso_tab[c("female", "male"), ])
min(ct_iso$expected)
ct_iso
```

H0: whether a researcher is connected is independent of gender.

HA: the two are not independent.

59% of women are isolated compared with 49% of men ($\chi^2$ = 5.63, df = 1, p = 0.018), so we reject H0. The article age check at the end of this section shows that this gap comes from women's articles being newer.

```{r rq5-degree, fig.width=5, fig.height=3.5, fig.cap="In-degree distribution on a log-log scale."}
dd = degree_distribution(g, mode = "in")
k = which(dd > 0) - 1
k = k[k > 0]
plot(k, dd[k + 1], log = "xy", pch = 19,
     xlab = "In-degree (log)", ylab = "Proportion of nodes (log)")
```

In-degree is heavily skewed (median 0, maximum 61) and falls roughly on a line on the log-log plot, the shape of a Barabasi-Albert graph where a few hubs get most of the links. The cluster of points around 30 is explained below.

```{r rq5-clique}
hub_ids = V(g)$name[degree(g, mode = "in") >= 28 & degree(g, mode = "out") >= 28]
clique  = induced_subgraph(g, hub_ids)
c(people = vcount(clique), edges = ecount(clique),
  share_of_all_edges = round(ecount(clique) / ecount(g), 2))
table(V(clique)$gender_group)
```

These 31 people form a complete clique (930 of 930 possible links, 42% of all links). This almost certainly comes from one navbox template rather than editors linking people one by one. 30 of the 31 are men, so we report centrality both with and without the links inside the clique.

## Centrality

In-degree counts how many researchers' articles link to a person. PageRank also weights each link by how important the linking article is, with a random surfer who follows a link 80% of the time ($\alpha$ = 0.8). PageRank scores are multiplied by the number of nodes, so 1 is average.

```{r rq5-centrality}
# remove only the links between clique members
g_nc = delete_edges(g, E(g)[V(g)[hub_ids] %->% V(g)[hub_ids]])

cent = data.frame(title  = V(g)$title,
                  gender = V(g)$gender_group,
                  era    = V(g)$era,
                  indeg  = degree(g, mode = "in"),
                  pr     = page_rank(g,    damping = 0.8)$vector * vcount(g),
                  pr_nc  = page_rank(g_nc, damping = 0.8)$vector * vcount(g))
cor(cent$indeg, cent$pr, method = "spearman")

# top 10 by PageRank, with and without the clique
top = cent[order(-cent$pr), c("title", "gender", "indeg", "pr")][1:10, ]
top$pr = round(top$pr, 1)
top$no_clique_top10 = cent$title[order(-cent$pr_nc)][1:10]
rownames(top) = NULL
top
```

The two measures rank people almost identically (Spearman's rho = 0.98). The top 10 are the field's founders and best-known deep learning researchers, and Fei-Fei Li is the only woman. Hinton, Minsky and McCarthy stay at the top without the clique, while Bengio, LeCun, Ng and Li drop out, so much of their centrality came from the template. Janet Kolodner enters the no-clique top 10 with only 3 incoming links because her page and Katia Sycara's link only to each other. The random surfer gets stuck in that loop, which is why we read PageRank alongside in-degree.

```{r rq5-gender}
mf = cent[cent$gender != "other/non-binary", ]

# share of women among the 50 most central people (dataset average: 24%)
sapply(c(full = "pr", no_clique = "pr_nc"),
       function(v) mean(cent$gender[order(-cent[[v]])][1:50] == "female"))

top10    = mf$pr    >= quantile(mf$pr,    0.9)
top10_nc = mf$pr_nc >= quantile(mf$pr_nc, 0.9)
round(prop.table(table(mf$gender, top10), 1), 3)

ct = chisq.test(table(mf$gender, top10))
min(ct$expected)
ct
chisq.test(table(mf$gender, top10_nc))$p.value
```

Centrality is too skewed to compare means, so we test whether gender is associated with being in the top 10% by PageRank.

H0: being in the top 10% by PageRank is independent of gender.

HA: the two are not independent.

Only 3.7% of women are in the top 10%, compared with 12.0% of men ($\chi^2$ = 11.9, df = 1, p < 0.001), and the result is the same without the clique (p < 0.001). Women are 24% of the dataset but only 14% of the 50 most central people. The gender gap is not caused by the navbox template.

```{r rq5-era}
ke = mf[mf$era != "unknown", ]
round(xtabs(pr_nc ~ era + gender, ke) / xtabs(~ era + gender, ke), 2)

# group sizes in the youngest era
table(ke$gender[ke$era == "1990+"])
```

Women's mean PageRank is below men's in every era. Women reach about 74% of men's level for those born before 1950 and about 86% for 1950-89, so the gap narrows but has not closed. The 1990+ group has only 2 women, which is too few to compare.

## Network plots

Women are orange and men are purple, which colour-blind readers can still tell apart. Node size shows PageRank on the full graph.

```{r rq5-plot-setup}
pr_full = page_rank(g, damping = 0.8)$vector
gender_col = c(female = "#E66100", male = "#5D3A9B", "other/non-binary" = "grey60")
dir.create("figures", showWarnings = FALSE)
```

The full plot shows the largest component (365 people, 97% of links), because the 474 isolated researchers would only add loose dots. We label the 8 most central women, since the top men all sit in the clique and their labels would overlap.

```{r rq5-plot, fig.width=9, fig.height=9, fig.cap="Largest connected component. Node size shows PageRank, colour shows gender."}
draw_full = function() {
  set.seed(3020)
  pr_l  = pr_full[V(lcc)$name]
  women = V(lcc)$name[V(lcc)$gender_group == "female"]
  top_w = names(sort(pr_l[women], decreasing = TRUE))[1:8]
  plot(lcc, layout = layout_with_fr(lcc),
       vertex.color = gender_col[V(lcc)$gender_group],
       vertex.size  = 1 + 7 * sqrt(pr_l / max(pr_full)),
       vertex.frame.color = NA,
       vertex.label = ifelse(V(lcc)$name %in% top_w, V(lcc)$title, NA),
       vertex.label.cex = 0.65, vertex.label.color = "black",
       vertex.label.dist = 0.8, vertex.label.font = 2,
       edge.arrow.size = 0.1, edge.color = adjustcolor("grey50", 0.3),
       main = "Largest connected component of AI researcher links (365 people)")
  legend("bottomleft", legend = c("Woman", "Man"), pch = 19, bty = "n",
         col = gender_col[c("female", "male")], pt.cex = 1.5)
}
draw_full()

png("figures/rq5_network_full.png", width = 1800, height = 1800, res = 200)
draw_full()
invisible(dev.off())
```

The dense centre is the clique, and apart from Fei-Fei Li no woman sits there. `r sum(V(lcc)$gender_group == "female")` of the `r vcount(lcc)` people in this component are women (`r round(100 * mean(V(lcc)$gender_group == "female"))`%), compared with 24% of the whole dataset.

For the subgraph we zoom in on the 40 researchers with the highest PageRank, because RQ5 asks who is central. Template links between clique members are light grey and links written by editors are dark grey. A circle layout keeps the clique from collapsing into a blob.

```{r rq5-subgraph, fig.width=9, fig.height=9, fig.cap="The 40 researchers with the highest PageRank. Light grey links come from the navbox template."}
top40 = names(sort(pr_full, decreasing = TRUE))[1:40]
sub   = induced_subgraph(g, top40)

# clique members first, then everyone else
ord = c(intersect(top40, hub_ids), setdiff(top40, hub_ids))
sub = permute(sub, match(V(sub)$name, ord))

# a link between two clique members comes from the navbox
from_to = ends(sub, E(sub))
is_nav  = from_to[, 1] %in% hub_ids & from_to[, 2] %in% hub_ids

lay_c = layout_in_circle(sub)
ang   = atan2(lay_c[, 2], lay_c[, 1])

draw_sub = function() {
  par(mar = c(1, 1, 3, 1))
  plot(sub, layout = lay_c, rescale = FALSE, xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5),
       vertex.color = gender_col[V(sub)$gender_group],
       vertex.size  = 3 + 8 * sqrt(pr_full[V(sub)$name] / max(pr_full)),
       vertex.frame.color = NA,
       vertex.label = V(sub)$title, vertex.label.cex = 0.6,
       vertex.label.color = ifelse(V(sub)$gender_group == "female", "#B34700", "black"),
       vertex.label.font = ifelse(V(sub)$gender_group == "female", 2, 1),
       vertex.label.dist = 1.6, vertex.label.degree = -ang,
       edge.arrow.size = 0.2, edge.width = ifelse(is_nav, 0.5, 1.5),
       edge.color = ifelse(is_nav, adjustcolor("grey70", 0.35), "grey15"),
       main = "Top 40 researchers by PageRank")
  legend("bottomleft", bty = "n", cex = 0.8,
         legend = c("Woman", "Man", "Navbox (template) link", "Other link"),
         col = c(gender_col[c("female", "male")], "grey70", "grey25"),
         pch = c(19, 19, NA, NA), lty = c(NA, NA, 1, 1), lwd = 2, pt.cex = 1.5)
}
draw_sub()

png("figures/rq5_top40.png", width = 1800, height = 1800, res = 200)
draw_sub()
invisible(dev.off())
```

```{r rq5-subgraph-text, include=FALSE}
women40 = V(sub)$title[V(sub)$gender_group == "female"]
```

Only `r length(women40)` of the 40 are women (`r paste(women40, collapse = ", ")`). 31 of the 40 are the clique, joined by `r sum(is_nav)` template links, and only `r sum(!is_nav)` editor-written links connect the other 9. Fei-Fei Li is there through the template and Daphne Koller through a few editor-written links. Kolodner and Sycara are only there because of the rank sink, so just 2 of the 40 (5%) are women with a real place in the network.

## Limitations

An article can only collect links after it is written, so we check whether article age explains the gender gap.

```{r rq5-article-age}
net = data.frame(gender   = V(g)$gender_group,
                 created  = people$created_year[match(V(g)$name, people$qid)],
                 indeg    = degree(g, mode = "in"),
                 isolated = degree(g) == 0,
                 pr       = page_rank(g, damping = 0.8)$vector * vcount(g))
net = net[net$gender != "other/non-binary", ]
net$top10 = net$pr >= quantile(net$pr, 0.9)

cor(net$created, net$indeg, method = "spearman")
tapply(net$created, net$gender, median)

# repeat both gender tests separately for older and newer articles
net$period = ifelse(net$created < 2015, "before 2015", "2015 onwards")
res = NULL
for (p in c("before 2015", "2015 onwards")) {
  s = net[net$period == p, ]
  ct_iso = chisq.test(table(s$gender, s$isolated))
  ct_top = chisq.test(table(s$gender, s$top10))
  res = rbind(res, data.frame(
    period         = p,
    women_isolated = round(mean(s$isolated[s$gender == "female"]), 2),
    men_isolated   = round(mean(s$isolated[s$gender == "male"]), 2),
    p_isolated     = round(ct_iso$p.value, 3),
    women_top10    = round(mean(s$top10[s$gender == "female"]), 3),
    men_top10      = round(mean(s$top10[s$gender == "male"]), 3),
    p_top10        = round(ct_top$p.value, 3),
    min_expected   = round(min(ct_iso$expected, ct_top$expected), 1)))
}
res
```

Newer articles get fewer links (Spearman's rho = -0.38), and women's articles are newer (median 2019 vs 2014 for men). Within each period women and men are equally likely to be isolated (p = 0.58 and 0.56), so the isolation gap is explained by article age. The top 10% gap is not. There is no difference for articles created before 2015 (14% vs 16%, p = 0.97), but for articles from 2015 onwards only 1% of women reach the top 10% compared with 8% of men (p = 0.003). Part of this is the navbox, which includes 14 of the 26 recent men in the top 10% and no recent women, but the gap remains without it (1% vs 6%, p = 0.03).

Other limitations: the network only includes people Wikidata tags as AI researchers with an English Wikipedia article, so links to anyone else are lost and part of the 52% isolation comes from this boundary. The links were collected once, and a request that failed three times was skipped without being logged. An edge means an editor mentioned someone, not that they collaborated, and edges are unweighted. PageRank depends on $\alpha$ = 0.8 and can be inflated by rank sinks. Gender comes from Wikidata and is mostly binary, so the 2 other/non-binary researchers were left out of the tests.

## Summary

Women are less central than men in the network of AI biographies. Only 3.7% of women are in the top 10% by PageRank compared with 12.0% of men, and only 2 of the 40 most central researchers are women with a real position in the network. This is not caused by the navbox template. The gap is concentrated in articles created from 2015 onwards, while older articles show no difference, and women's higher isolation is explained by their articles being newer.
